---
output: github_document
bibliography: /Users/zbigniewkarpinski/ReferencesMaster.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stat.alloc

<!-- badges: start -->

<!-- badges: end -->

stat.alloc is an R package with a set of functions which allow for calculating reference allocations for a given (observed) status allocation and use these reference allocations to estimate models in a family of status allocation models proposed in a series of publications by Zbigniew Karpiński and John Skvoretz [see @karpinski2023 for example]. Consider the original publications for details concerning the models.

## Installation

You can install the development version of stat.alloc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zbig-karp/stat.alloc")
```

## Example

This is a basic example which shows you how to calculate **meritocratic** and **lottery** allocations from a given (observed) status allocation [check @krauze1985 and @karpinski2023 for formal definitions of the various types of status allocation].

The code below defines an example of a status allocation table. The table is reproduced from a seminal paper by @krauze1985. It shows counts of individuals from a given educational category (row of the matrix) who end up in a given occupational status (column of the matrix):

```{r table-1}
ks1985 <- matrix(
  c(135, 52, 58, 12, 6, 
    27, 56, 126, 26, 7, 
    13, 51, 183, 98, 65, 
    3, 8, 30, 23, 21), 
  ncol = 4, dimnames = list(rownames = paste0("$E_", 1:5, "$"), 
                            colnames = paste0("$O_", 1:4, "$")))
```

Two **reference allocations** can be proposed for a given status allocation: a **meritocratic** allocation and a **lottery allocation**. Under the former, individuals are assigned to destination statuses based on their merits, so that those with the highest merits are the first to be assigned to the highest status. As a result, under perfectly meritocratic allocation, it is never the case that a person with lower merits ends up in higher destination status than a person with higher merits. Formally, the result of the fully meritocratic status allocation is defined as follows: 

$$
m_{gh} = \min{\left(a_{g+} - \sum_{i=0}^{h-1}m_{gi},\,a_{+h} - \sum_{j=0}^{i-1} m_{jh}\right)},
$$ 

where $m_{gh}$ is the number of individuals from education category $g$ who end up in occupational status $h$, $a_{g+}$ is the $g$^th^ row marginal and $a_{+h}$ is the $h$th column marginal. Further, we set $m_{g0} = m_{0h} = 0$ for simplicity of expression. Allocation by this procedure insures that it is never the case that a person with lower credentials is found in a higher destination status than a person with higher credentials.

```{r example}
library(stat.alloc)

mar(ks1985)

```

# References
