---
output: github_document
bibliography: /Users/zbigniewkarpinski/ReferencesMaster.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stat.alloc

<!-- badges: start -->

<!-- badges: end -->

stat.alloc is an R package with a set of functions which allow for calculating reference allocations for a given (observed) status allocation and use these reference allocations to estimate models in a family of status allocation models proposed in a series of publications by Zbigniew Karpiński and John Skvoretz [see @karpinski2023 for example]. Consider the original publications for details concerning the models.

## Installation

You can install the development version of stat.alloc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zbig-karp/stat.alloc")
```

## Example

This example walks through a calculation of **meritocratic** and **lottery** allocations from a given (observed) status allocation [check @krauze1985 and @karpinski2023 for formal definitions of the various types of status allocation] and uses the result to estimate a formal model of status allocation.

We first set up a status allocation table. It shows counts of individuals from a given educational category (row of the matrix) who end up in a given occupational status [column of the matrix; see @krauze1985 for the source of the data]:

```{r table-1}
ks1985 <- matrix(
  c(135, 52, 58, 12, 6, 
    27, 56, 126, 26, 7, 
    13, 51, 183, 98, 65, 
    3, 8, 30, 23, 21), 
  ncol = 4, dimnames = list(rownames = paste0("E", 1:5), 
                            colnames = paste0("O", 1:4)))
```

Function `mar()` is used to calculate meritocratic and lottery allocations from a given one:

```{r example}
library(stat.alloc)

mar(ks1985)
```

# References
