---
output: github_document
bibliography: /Users/zbigniewkarpinski/ReferencesMaster.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(tidyverse)
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(digits = 2)
```

# stat.alloc

<!-- badges: start -->

<!-- badges: end -->

stat.alloc is an R package with a set of functions which allow for calculating reference allocations for a given (observed) status allocation and use these reference allocations to estimate models in a family of status allocation models proposed in a series of publications by Zbigniew Karpiński and John Skvoretz [see @karpinski2023 for example]. Consider the original publications for details concerning the models.

## Installation

You can install the development version of stat.alloc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zbig-karp/stat.alloc")
```

## Example

This example walks through a calculation of **meritocratic** and **lottery** allocations from a given (observed) status allocation [check @krauze1985 and @karpinski2023 for formal definitions of the various types of status allocation] and uses the result to estimate a formal model of status allocation.

We first set up a status allocation table. It shows counts of individuals from a given educational category (row of the matrix) who end up in a given occupational status [column of the matrix; see @krauze1985 for the source of the data]:

```{r table-1}
ks1985 <- matrix(
  c(135, 52, 58, 12, 6, 
    27, 56, 126, 26, 7, 
    13, 51, 183, 98, 65, 
    3, 8, 30, 23, 21), 
  ncol = 4, dimnames = list(rownames = paste0("E", 1:5), 
                            colnames = paste0("O", 1:4)))
```

Function `mar()` is used to calculate meritocratic and lottery allocations from a given one:

```{r example-1}
library(stat.alloc)

mar(ks1985)
```

## Constant mixing models

Function `cmm_mde()` estimates a model which treats the observed status allocation as a mixture of two "pure" status allocation. It returns a list of three elements:

- mixing coefficient, or the proportion of meritocracy in the mix,
- model-predicted status allocation,
- the index of dissimilarity between the observed and model-predicted allocations

```{r example-2}
cmm_mde(dat = mar(ks1985))
```

`cmm_mde()` estimates the mixing coefficient by searching for a value of the coefficient which minimises the distance (or the Frobenius norm of the difference) between the observed and model-predicted status allocations. An alternative, a function called `cmm_mle`, estimates the coefficient by means of maximum likelihood. 

```{r example-3}
cmm_mle(dat = mar(ks1985))
```

As we can see, the maximum-likelihood estimate is slightly lower than the minimum-distance one and it has a considerably larger standard error. In terms of fit, as measured with the dissimilarity index, the two estimations are nearly identical, but their predicted counts differ slightly. The index of dissimilarity between the two model-predicted allocations is:

```{r}
pred_mde <- prop.table(cmm_mde(dat = mar(ks1985))[[2]])
pred_mle <- prop.table(cmm_mle(dat = mar(ks1985))[[2]])
sum(abs(pred_mde - pred_mle))/2
```

Thus, `r round(1000 * sum(abs(pred_mde - pred_mle))/2)` cases (individuals) out of 1,000 have to be re-allocated to equalise the two matrices. It's not much in absolute terms, but it still shows that the two predictions differ.

## Differential mixing models

So far, the mixing coefficient was assumed to be constant across all origin categories. It is possible to relax this assumption and allow each origin category to have its own mixing coefficient. Function `dmm_mde()` fits a **differential mixing model** using the minimum distance estimation routine:

```{r example-4}
dmm_mde(dat = mar(ks1985))
```

The object returned by `dmm_mde` has a structure similar to those returned by `cmm_mde` or `cmm_mle`. A primary difference is that `dmm_mde` returns a list of four, rather than three, elements, the additional element being a named numeric vector of so-called adjustment proportions, or a special type of parameter introduced into the model to fit the column marginals [see @karpinski2023 for more details concerning the adjustment proportions].

There is also a function which fits the differential model by means of maximum likelihood estimation, `dmm_mle()`. 

```{r example-5}
dmm_mle(dat = mar(ks1985))
```

# References
