---
output: github_document
bibliography: /Users/zbigniewkarpinski/ReferencesMaster.bib
nocite: |
  @krauze1985
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stat.alloc

<!-- badges: start -->
<!-- badges: end -->

stat.alloc is an R package with a set of functions which allow for calculating reference allocations for a given (observed) status allocation and use these reference allocations to estimate models in a family of status allocation models proposed in a series of publications by Zbigniew Karpiński and John Skvoretz [see @karpinski2023 for example]. Consider the original publications for details concerning the models.

## Installation

You can install the development version of stat.alloc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zbig-karp/stat.alloc")
```

## Example

This is a basic example which shows you how to calculate meritocratic and lottery allocation for a given (observed) status allocation:

```{r example}
library(stat.alloc)

# An example of a status allocation table reproduced from a seminal paper by Krauze and Słomczyński (1985). It shows counts of individual from a given educational category (row of the matrix) who end up in a given occupational status (column of the matrix).

ks1985 <- matrix(
  c(135, 52, 58, 12, 6, 
    27, 56, 126, 26, 7, 
    13, 51, 183, 98, 65, 
    3, 8, 30, 23, 21), 
  ncol = 4, dimnames = list(
    rownames = paste0("$E_", 1:5, "$"), 
    colnames = paste0("$O_", 1:4, "$")))

# Reference status allocations 

mar(ks1985)
```

# References
